---
title: "POLI 572B"
author: "Michael Weaver"
date: "April 10, 2026"
widgets: [bootstrap]
output: 
  revealjs::revealjs_presentation:
    theme: white
    highlight: haddock
---

```{r setup, include = F}
require(knitr)
require(magrittr)
require(kableExtra)
require(ggplot2)
require(grid)
require(data.table)
require(UsingR)
require(lfe)

options("kableExtra.html.bsTable" = T)
```


<style type="text/css">
  .reveal h2,h3,h4,h5,h6 {
    text-align: left;
  }
  .reveal p {
    text-align: left;
  }
  .reveal ul {
    display: block;
  }
  .reveal ol {
    display: block;
  }
  .table-hover > tbody > tr:hover { 
  background-color: #696969;
  }
</style>


## Outline

### Natural Experiments

- Motivation/Definition
- Challenges
- IV
- RDD
- Qualitative Evidence
- Balance Tests
- Papers



# Natural Experiments

---

### What do we mean by natural experiment?

An **observational study** where causal inference comes from the design that draws on randomization of treatment.

- no manipulation of treatment by researcher
- random or "as-if" random assignment takes place in the "real world"


**experiment**: because it has **randomization**

**natural**: because it occurs without researcher intervention in real-world

---

### Why?

1. solution to confounding

2. no complex estimands (hopefully!), simple and transparent statistical models
  
  - fewer worries about common support, interpolation/extrapolation


## Compare and contrast

> Piccardi et al use simple $t$ tests (in a regression context) to compare differences in means

> "The power of multiple regression analysis is that it allows us to do in non-experimental environments what natural scientists are able to do in a controlled laboratory setting: keep other factors fixed" (Wooldridge 2009: 77)

## Design vs. Model-based inferences

A matter of **degree**

Statistical evidence for causality combines **observed data** and a **mathematical model** of the world.

>- mathematical model makes inferences about reality
>- model always involves assumptions about how observed data are generated
>- can almost always apply the model, even if it is wrong

## Design vs. Model-based inferences

Causal evidence varies in terms of complexity of math/restrictiveness of assumptions: a matter of **degree**

- **Model-based** inferences about causality involve many choices in complex statistical models with many difficult-to-assess assumptions

- **Design-based** inferences about causality use carefully controlled comparisons with simple models and transparent assumptions

---

### Challenges

1. Finding Natural Experiments
    - relevance of intervention; relevance of randomly assigned group; where is there actually randomness?
    
    - random vs as-if random: known random process (lottery), something that is "as-if" random... (in what way)
    
2. Arguing for Randomness

- how do we know its random? lotteries maybe (but you need to know whether rules followed... protocol correct)
- qualitative evidence
- balance tests... shouldn't see pre-treatment differences (in what, though?)

3. How simple is the design?

- Isolating the random exposure sometimes involves complicated models as well... not always so simple to understand


---

### Varieties:

1. "Standard natural experiments": treatment $D$ is randomized for some units

2. Instrumental Variables: some instrument $Z$ is randomized and affects treatment $D$

3. Regression Discontinuity: treatment assignment of $D$  takes place at cutoff $c$, random close to $c$

# Instrumental Variables


---

### IV 

- basically extends CACE: continuous treatments/instruments; random assignment conditional on other variables

assumes random $Z$, exclusion restriction, monotonicity (no defiers)

- for "compliers"... harder to define with continuous treatment
- can add controls

2sls also assumes linearity; if treatment is continuous... which values of $D$ are contributing to effect estimate?


---

### IV Example: QJPS paper


equation... 

explain the logic


---

### IV Issues 

PAPER LINK

weak instruments => big standard errors

fix your standard errors: main suggested tests

---

### IV Issues 

weak instruments => bias amplification (exclusion restriction problems)

zero first stage; affect possible excluded paths?; 

---

reduced form tests

# RDD

--- 

### Regression Discontinuity

Classic example: 

Effect of criminal vs. non-criminal politicians winning elections... compare places with criminal candidate as winner or runner up.

there is a continuous variable (forcing variable): margin of victory (criminal - non-criminal vote share)

When margin of victory reaches 0 (vs next candidate) move from untreated to treated.

---

FIGURE 1 from titiunik


---

Move from all untreated to all treated (sharp RD; fuzzy RD does this with IV or reduced form)

E[Y_i(1) - Y_i(0) | X = c]


---

fig 2 from titiunik

---

Key assumption:

- continuity at cutoff conditional expectation functions of Y(1), Y(0) across forcing variable (vote share)
- it can be changing... but expected to change SMOOTHLY at the cutoff


E[Y_i(1)|X] E[Y_i(0) | x]

---

How do we get the CEF further out?

fpci; figure from titiunik


---

Only need to estimate it at the cutoff...

=> local polynomial regression; estimate the slope of E[Y(0) | x] AT x = c; vice versa


---

CHoices

- polynomial order
- kernel
- bandwidth

---

- choosing polynomial order

---

- choosing kernel

--- 


choosing bandwidth


---

standard errors:

- correct choice is bias corrected (see what is recommended)


---

checks:

pre-treat covariates: run with their own RDD... do you observe discontinuities... if so... possibly not continuous

density tests (sorting test): 

"donut": leave out cases right on the cutoff (possible sorting)


# Assessing Random Assignment

---

### Balance Tests

Logic: if random, we should not see significant differences in other attributes

- t-tests/regressions; ks tests
- classically, hypothesis tests. should so significant differences at rate of alpha
- equivalence tests are better: what range of differences can you exclude?
- pre-treatment covariates

---

But which covariates?

Qualitative evidence...


---


Qualitative evidence: causal process observation of treatment assignment.

How is treatment actually assigned in reality?

Ask about units being treated/actors contrlling treatment...


information

incentives

capacities


---


Papers:

Discussion of Kocher and Monteiro vs Ferwerda and Miller.


---

Papers: QJPS... what is the design, what are the tests doing...












## Natural Experiments:

### Regression Discontinuity

- "Treatment" occurs for cases on one side of threshold, not to cases on other side
- Near the "threshold", cases are arbitrarily assigned.
    - treatment can be for ALL cases on the "Treated" side (sharp RD) 
    - treatment can increase at the threshold ("fuzzy" RD) (e.g. Mo 2018)
   
   
## Natural Experiments

```{r, echo = F, message=F}
n = 5000
x = runif(n, -1,1)
y = 3 + 1*x + 2*(x>0) + rnorm(n)

d = data.frame(x,y, d = x>0)
ggplot(d, aes(x = x, y = y, group = d)) + 
  geom_point() + 
  geom_smooth() +
  geom_vline(xintercept = 0)
  theme_bw()

```
    
## Natural Experiments:

### Regression Discontinuity

Two approaches:

- Potential outcomes of $Y$ are approximately **linear** near the threshold; treatment shifts outcomes
    - so we (flexibly) model $Y$ on either side of the cut-off
- Potential outcomes of $Y$ are random near the cut-off
    - difference in means near the cutoff

## Natural Experiments:

### Regression Discontinuity

Decisions:

- model outcome around cutoff
    - bias/variance trade-off
    - how do you model it? linear? quadratic? cubic? quartic? local linear?
    - how do you choose the "bandwidth" around the cutoff to use?
    - typically recommend local linear, automatic choice of 'optimal bandwidth'

- difference in means at cutoff
    - how do you choose the "bandwidth" around the cutoff to use?

## Natural Experiments:

### Regression Discontinuity

Assumptions:

- assignment to treatment at the threshold is random
- AND
    - either we have correctly modeled potential outcomes of $Y$ on either side of cutoff 
    - or we have used "bandwidth" around cutoff within which treatment is as-if random assigned (difference in means)

## Natural Experiments:

### Instrumental Variables

Follows from Wald estimator for non-compliance:

- Instrumental Variables Least Squares uses variance "explained" in $X$ by  randomly induced $Z$
    - $X$ is "treatment", $Z$ is "random assignment to treatment"
    - easy to understand when $Z$ is binary
    - when $Z$ is continuous, it is more complicated (and depends on assumptions of linearity)

## Natural Experiments:

### Instrumental Variables

Decisions

- is variation induced in $X$ by $Z$ linear?
- does $Z$ actually predict change in $X$?
    - "weak instruments problem": identical to when there are very few compliers in an experiment: inference becomes very hard
    - typically do $F$ test of regression of $Y$ on $X$; $F$ statistic $> 10$.


## Natural Experiments:

### Instrumental Variables

Assumptions:

1. $Z$ is randomly assigned
2. Exclusion Restriction
    - no other causal path from $Z$ to $Y$, other than $X$
3. Monotonicity
    - increasing $Z$ uniformly increases or decreases values of $X$

## Natural Experiments:

### Choices

- Sometimes natural experiments only random conditional on some other variables (Nellis, et al.)
    - What is the correct functional form for the conditioning variables?
- Standard Errors:
    - What is the unit at which treatment is assigned?
    - Cluster/Randomization inference

## Natural Experiments:

### Limitations

- Effects are "local" or for "compliers"
    - like in non-compliance, we find subset of cases that are randomly induced into treatment/responsive
    - cases may be different

- Causal exposures may be different
    - cause applied at cutoff might be different than elsewhere
    - cause applied at random might be different (e.g. rainfall as instrument for economic growth)
    
## Natural Experiments:

### What to do:

- Show robustness to choices of analysis
- Evidence that randomization took place
    - statistical tests of implication
    - qualitative evidence: "causal process observation"