---
title: "POLI 572B"
author: "Michael Weaver"
date: "March 7, 2022"
widgets: [bootstrap]
output: 
  revealjs::revealjs_presentation:
    theme: white
    highlight: haddock
---

```{r setup, include = F}
require(knitr)
require(magrittr)
require(kableExtra)
require(ggplot2)
require(grid)
require(data.table)
require(UsingR)
options("kableExtra.html.bsTable" = T)

acs_data = fread("acs_lawmed.csv")
acs_data[, FEMALE := SEX]
acs_data[, MARST_f := as.factor(MARST)]
set.seed(1234)
acs_data = acs_data[sample(1:nrow(acs_data), size = 10000)]
acs_data[, MALE := abs(FEMALE - 1)]
acs_data[, sex := as.factor(ifelse(FEMALE == 1, 'Female', 'Male'))]
```

<style type="text/css">
  .reveal h2,h3,h4,h5,h6 {
    text-align: left;
  }
  .reveal p {
    text-align: left;
  }
  .reveal ul {
    display: block;
  }
  .reveal ol {
    display: block;
  }
  .table-hover > tbody > tr:hover { 
  background-color: #696969;
  color: #FFFFFF;
  }
</style>

# Inference with Least Squares

--- 

### Objectives

**Recap**

- Least Squares as a predictive algorithm


**Inference**

- Least Squares as an **estimator** of some unknown **parameters**

---

### Today

Least Squares 

- statistical model for inference
    - what is the statistical model? (like, e.g. Neyman Causal Model)
    - what are the assumptions we need for...
        - unbiasedness of estimator
        - deriving uncertainty of estimate
        
# Recap

---

### Least Squares for Prediction

So far, we...

- ...start with a vector $\mathbf{Y} = (Y_1 \ldots Y_i \ldots Y_n)$

- ... want to produce a vector of **predicted values** $\mathbf{\widehat{Y}} = (\widehat{Y_1} \ldots \widehat{Y_i} \ldots \widehat{Y_n})$

- .... defined as $\mathbf{\widehat{Y}} = \mathbf{X}\mathbf{\beta}$:

    - where $\mathbf{X}$ is a matrix of predictors (for the mean, just a vector of $1$s) and $\mathbf{\beta}$ is a vector of linear coefficients that are multiplied by $\mathbf{X}$ such that $\widehat{Y}_i = b_0x_0 + b_1x_1 + \ldots + b_px_p$

---

### Least Squares for Prediction


Then, we...

- ... choose $\beta$ such that $\mathbf{\widehat{Y}}$ have **minimum (euclidean) distance** to $\mathbf{Y}$:

    - minimize  $\mathbf{Y} - \mathbf{\widehat{Y}} = \mathbf{e}$; where $\mathbf{e}$ is a vector of **residuals** (prediction errors) $(Y_1 - \hat{Y_1}, \ldots , Y_i - \hat{Y_i}, \ldots, Y_n - \hat{Y}_n)$
  
-  $\mathbf{e}$ is minimized when $\mathbf{e}'\mathbf{X} = \mathbf{0}$: residuals are **orthogonal** to $\mathbf{X}$.

This gave us $\mathbf{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$

---

### Least Squares for Prediction

This definition of least squares means that:

- residuals $\mathbf{e}$ are orthogonal to $\mathbf{X}$; so if there is an intercept, the mean of the residuals is $0$, and covariance of residuals and $\mathbf{X}$ is $0$.
    
- If we just fit an intercept (same value of $\widehat{Y}$ for all cases): what is the mean of the residuals $E(Y_i - \widehat{Y_i})$?


--- 

### Example

Let's say we want to get the **conditional expectation function** for yearly earnings of professionals as a function of hours worked, profession, gender, and age.

Using data from the American Community Survey, we can look at how hours worked is related to earnings for doctors and lawyers. 

- `UHRSWORK` (the usual hours worked per week)
- `FEMALE` (an indicator for female)
- `AGE` (in years)
- `LAW` (an indicator for being a lawyer)
- `INCEARN` (total annual income earned in dollars). 

---

### Example

Let's now find the (approximate) linear conditional expectation function of earnings, across hours worked, age, profession, and gender:

```{r}
ls = lm(INCEARN ~ UHRSWORK + AGE + LAW + FEMALE, acs_data)
```

---

### Example

```{r, echo = F}
summary(ls)
```

```{r, echo = F, include = F, message = F, warning=F}
residual_hrswork = lm(FEMALE ~ AGE + LAW + UHRSWORK, acs_data)$residuals
plot_data = rbind(acs_data[, list(INCEARN, AGE, LAW, FEMALE, UHRSWORK)], acs_data[, list(INCEARN, AGE, LAW, FEMALE, UHRSWORK)])
plot_data[, type := factor(rep(c('Observed', 'Residual'), each = nrow(acs_data)))]
plot_data[, FEMALE := as.numeric(FEMALE)]
plot_data[type == 'Residual', FEMALE := residual_hrswork]
```

How do we make sense of, e.g., the slope on `FEMALE`?

---

### Example

```{r, echo = F, warning=F, message=F}
ggplot(plot_data, aes(x = AGE, y= FEMALE)) + geom_point(alpha = 0.01) + geom_smooth(method = 'lm', colour = 'red', se = F) +  geom_smooth(method = 'loess', colour = 'blue', se = F) + theme_bw() + facet_wrap(~type) + ggtitle("Gender Residual on Age")
```

---

### Example

```{r, echo = F, warning=F, message=F}
ggplot(plot_data, aes(x = LAW, y= FEMALE)) + geom_point(alpha = 0.01)  + geom_smooth(method = 'lm', colour = 'red', se = F) + geom_smooth(method = 'loess', colour = 'blue', se = F) + theme_bw() + facet_wrap(~type) + ggtitle("Gender Residual on Profession")
```

---

### Example

```{r, echo = F, warning=F, message=F}
ggplot(plot_data, aes(x = UHRSWORK, y= FEMALE)) + geom_point(alpha = 0.01) + geom_smooth(method = 'lm', colour = 'red', se = F) + geom_smooth(method = 'loess', colour = 'blue', se = F) + theme_bw() + facet_wrap(~type)  + ggtitle("Gender Residual on Hours Worked")
```

---

### Example

```{r, echo = F, warning=F, message=F}
ggplot(plot_data, aes(x = FEMALE, y = INCEARN)) + geom_point(alpha = 0.01) + geom_smooth(method = 'lm', colour = 'red', se = F) + theme_bw() + facet_wrap(~type)  + ggtitle("Earnings on Residual Gender")
```

---

Looking at residuals, gender and hours worked have *non-linear* relationship. How can we turn this *non-linear* relationship into a *linear* relationship?

Angrist and Pischke, *MHE*, pp. 77-80

>- transform continuous variables into dummies: dummies can **only** have a linear relationship
>- ensures that least squares correctly captures relationship between variables

# Inference

---

### Regression for Inference:

1. We want to interpret the $\beta$ vector causally.

- typically just one coefficient of causal interest: why? (think about DAGs)

2. There is a true $\beta_D$ ($\delta$) that is the effect of some $D$ on $Y$.

- either a true linear effect, or weighted average of non-linear effects

3. $\beta_D$ is an unknown parameter. 

4. To **estimate** $\beta_D$, need a **statistical model** for least squares

---

### A Statistical Model

As with experiments (e.g. the Neyman causal model), **statistical models** 

- treat observed data as **realizations** of **random variables** (some chance process)
- with some assumptions...
- permit us to connect what we observe to **unobserved parameters**

Need to clarify what work is being done by

- data
- regression mathematics
- assumptions of statistical model

---

### An example: Hooke's Law

> Hooke's law is a law of physics that states that the force ($D$) needed to extend or compress a spring by some distance $y$ increases **linearly** with respect to the distance. 

Thus, for a given weight $D_i$ on a spring, the length $Y_i$ is determined by:

$Y_i = \alpha + \beta D_i + \epsilon_i$ for $i$ in $1 \ldots n$



---

### An example: Hooke's Law

$Y_i = \alpha + \beta D_i + \epsilon_i$ for $i$ in $1 \ldots n$

This is a **statistical** model:

- $\alpha$ and $\beta$ are unknown **parameters** (thus, greek letters, not $a$ and $b$). They are **fixed**.
    - $\alpha$ is the length of spring in absence of force
    - $\beta$ is the change in length of spring with unit of force
- $\epsilon_i$ is an unknown **random error** (a **random variable**). Each $\epsilon_i$ is...
    - **independent** of $D_i$  
    - each drawn from the **same distribution** with mean $0$ and variance $\sigma^2$ (another parameter)

We use observed data to **estimate** the **parameters** $\alpha, \beta, \sigma^2$

---

### In groups:

Imagine we have a spring whose length is 3 cm. With each Newton of force applied, the spring increases by 0.5 cm.

Using these facts and the statistical model described above: create a **response schedule**/**potential outcomes** table corresponding to the spring under loads of $0,1,2,3,5,8$ Newtons.

$Y_i(D_i = 0), Y_i(D_i = 1), Y_i(D_i = 2), \ldots$

---

### The Regression Model: 

The statistical model assumes that this linear equation describes the process generating the data.

$$Y_i = \alpha + \beta D_i + \epsilon_i$$

**deterministic** (not random)
$\alpha + \beta D_i$

**stochastic** (random error)
$\epsilon_i$


---

### The Regression Model:

$\alpha$: parameter for the intercept (fixed for all cases)

$\beta$: parameter for the slope (fixed for all cases)

$D_i$: value of independent variable (input into the equation)

$Y_i$: value of dependent variable (determined by the equation)

$\epsilon_i$: random error for observation $i$

- each $\epsilon_i$ is realization of same random variable with mean $0$ and variance $\sigma^2$. 
- **independent** of each other *and* $D_i$

---

```{r, echo = T}
d = rnorm(100); y = 3 + d * 0.5
```

```{r, echo = F}
plot(d, y, main = bquote(Y[i] == alpha + beta * D[i])); abline(a = 3, b = 0.5)
```

---

```{r, echo = T}
epsilon_i = rnorm(100); y_star = 3 + d * 0.5 + epsilon_i
```

```{r, echo = F}
plot(d, y_star, main = bquote(Y[i] == alpha + beta * d[i] + epsilon[i]))
abline(a = 3, b = 0.5)
```

---

In the context of Hooke's law and other physical laws may make sense to expect the effect $\beta_D$ to be linear and fixed for all observations.

Think back to experiments: in context of political science, does this make sense?

- e.g., is the effect of income on preferences for redistribution the same for all people? Linear?

---

### The Regression Model:

It is possible to think of regression estimates like this:

- $\beta_D$: is weighted average of non-linear effect of $D$ on $Y$ across different values of $D$
- $\beta_D$: can represent the average effect, even if there are heterogeneous unit causal effects

$Y_i = \alpha_i + \beta_{D_i} D_i + \nu_i$
$Y_i = \alpha_i + (\beta_{D} + (\beta_{D_i} - \beta_{D})) D_i + \nu_i$
$Y_i = \alpha_i + \beta_{D} D_i + (\nu_i + (\beta_{D_i} - \beta_{D})D_i)$

... if heterogeneous effects are **independent** of $D$

---

### The Regression Model: 

The more general **regression model** (where $X_i$ is a matrix of variables)

$$Y_i = \mathbf{X_i\beta} + \epsilon_i$$

can be **estimated** with least squares:

$$Y_i =  X_i\hat{\beta} + e_i$$

where residuals used to estimate variance of errors: $Var(e) = \widehat{\sigma^2}$

---

### The Regression Model:

**Differences from Least Squares Algorithm**

- we now have **estimates** ($\widehat{}$) of **parameters**
- we now have **unobserved errors** $\epsilon_i$ instead of **residuals** $e_i$.

Least Squares is the **Best Linear Unbiased Estimator** of our model parameters under 3 key assumptions.
  
- by unbiased we mean $\beta - E(\widehat{\beta}) = 0$
- "best" as in estimate has least variance (not so important in practice)

---

### Model Assumptions:

$(1)$. **Model** equation is correctly specified 

- correct functional form (e.g., linear, additive) for process by which variables in $X$ affect $Y$, and how variables in $X$ **affect each other**.
- all **relevant variables** affecting $Y$ (will define next week) are included in the model
- $Y_i =  \mathbf{X_i}\mathbf{\beta} + \epsilon_i$ is the true data-generating function

In groups: come up with 2 ways this assumption could be wrong in the context of this model: 

$Earn_i = b_0 + b_1 Hours_i + b_2 Age_i + b_3 Law_i + b_4 Female_i + \epsilon_i$

---

### Model Assumptions:

$(2.)$ $\epsilon_i$ are independent, identically distributed (i.i.d.)

- $E(\epsilon_i) = 0$ and variance of $\sigma^2$. 
- **homoskedastic**
    - errors $\epsilon$ all come from same distribution with given variance
- in contrast to **heteroskedastic**: 
    - errors come from different distributions with different variances.

In groups: come up with 2 ways this assumption could be wrong in the context of the Hooke's Law example.

---

### Model Assumptions:

$(3.)$ $\epsilon_i$ is **independent** of $X_i$: $X_i \perp \!\!\! \perp \epsilon_i$

- This means $X$ is **independent** of $\epsilon$ (recall dependence versus independence)
- **independence** is not **orthogonality**
    
In groups: come up with 2 ways this assumption could be wrong in the context of this model: 

$Earn_i = b_0 + b_1 Hours_i + b_2 Age_i + b_3 Law_i + b_4 Female_i + \epsilon_i$
    
---

### Model Assumptions:

Under these three assumptions, we call the model "ordinary least squares".

We cannot **empirically verify** these assumptions.

In particular, cannot verify assumption 3:

- When using least squares to **estimate** this model, we mathematically impose $e_i$ to be **orthogonal** to $x_i$ and $E(e_i) = 0$. This is a mathematical feature of regression, not proof that $X_i \perp \!\!\! \perp \epsilon_i$

--- 

### Model Assumptions:

If these assumptions are reasonable for the data we observe, then with large $n$, we can estimate our parameters $\beta, \sigma^2$ using least squares. 

And $\beta$ could have a causal interpretation.


# Estimating the Regression Model


---

### The Regression Model:

What work is done by

- the data?
- the math of regression?
- the assumptions?

---

### The Regression Model:

**Mathematical Assumptions** (least squares):

- **design matrix** has **full rank** (no linear dependence b/t columns)
- $n \geq p$

**Statistical Assumptions** (statistical model):

- $Y_i$ is generated by $\mathbf{X_i\beta} + \epsilon_i$
- $\epsilon_i$ are independent, identically distributed, with variance $\sigma^2$
- $X_i \perp \!\!\! \perp \epsilon_i$


---

### Estimation:

If $\mathbf{X}$ is $n \times p$ matrix: We want to estimate $p + 1$ parameters... (what are they?)

$\beta$ estimated by $\widehat{\beta}$ using least squares:

$$\widehat{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$$

- Note: $\widehat{\beta}$ is a random vector; each element is a **random variable**. **Why?**

---

$Y_i = X_i\beta + \epsilon_i$

$\epsilon_i$ is random variable.

If we were to repeat the drawing of $\epsilon$ again, we would get different estimates: $\widehat{\beta}$.


---

```{r, echo = T}
d = rnorm(100); y = 3 + d * 0.5
```

```{r, echo = F}
plot(d, y, main = bquote(Y[i] == alpha + beta * D[i])); abline(a = 3, b = 0.5)
```

---

```{r, echo = T}
epsilon_i = rnorm(100); y_star = 3 + d * 0.5 + epsilon_i
```

```{r, echo = F}
plot(d, y_star, main = bquote(Y[i] == alpha + beta * d[i] + epsilon[i]))
abline(a = 3, b = 0.5)
```

---

```{r, echo = F}
k = 1000
out = data.frame(alpha_hat = rep(NA, k), beta_hat =  rep(NA,k))

alpha = 3
beta = 0.5

for(i in 1:k){
  y_star = alpha + beta*d + rnorm(100)
  bs_m = lm(y_star ~ d)
  out[i,] = bs_m$coefficients
}

u = data.frame(d, y)

ggplot(u, aes(x = d, y=y)) + geom_point(alpha = 0) +
  xlab('d') + ylab('y') +  
  geom_abline(intercept = out$alpha_hat, slope = out$beta_hat, colour = 'red', linetype = 'dashed', alpha = 0.025) + geom_abline(intercept = alpha, slope = beta, colour = 'darkred') + theme_bw() + labs(title = 'Lines from 1000 draws of epsilons')

```

---
  
```{r, echo = F, warning=F, message=F}

ggplot(out, aes(x = beta_hat)) + geom_histogram() + theme_bw() + ggtitle("Sampling Distribution of Beta Hat")

```

---

### Regression and Random Variables

Compare this against the logic of experiments:

- in experiments, random assignment makes mean observed potential outcomes of $Y$ into random variables.
- in regression, what is the "chance process" generating $\epsilon$?
>- Doesn't really make a lot of sense; even recent textbooks hand-wave and say it isn't important to interpret this; the math just works
>- A big open question, because $\epsilon$ drives our variance estimates for $\widehat{\beta}$

---

### Estimation:

Under the statistical assumptions we made, $\widehat{\beta}$ is an unbiased estimator of $\beta$, conditional on $X$: $E(\widehat{\beta}|X) = \beta$
  
- "conditional on $X$" means taking the values of $X$ as fixed; not a sample from a population

--- 

### LS estimate is unbiased:

By assumption: $Y = \mathbf{X\beta} + \epsilon$, so:

$\widehat{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$

$\widehat{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'(\mathbf{X\beta} + \epsilon)$

$\widehat{\beta} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon$

$\widehat{\beta} = \mathbf{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon$

---

### LS estimate is unbiased:

$\widehat{\beta} = \mathbf{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon$

So:

$E(\widehat{\beta}|X) = E(\mathbf{\beta}|\mathbf{X}) + E((\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon|\mathbf{X})$

$E(\widehat{\beta}|X) = \mathbf{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'E(\epsilon|\mathbf{X})$

And because, by assumption, $X \perp \!\!\! \perp \epsilon$

$E(\widehat{\beta}|X) = \mathbf{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'E(\epsilon)$

And because, by assumption, $E(\epsilon) = 0$

$E(\widehat{\beta}|X) = \mathbf{\beta}$

---

### Variance of LS Estimator?

Thus, if our model assumptions are correct, $\widehat{\beta}$ is an unbiased estimate of causal parameter $\beta$

- $\widehat{\beta}$ is a random vector, as it depends on chance draws of $\epsilon$
- As $n \to \infty$, $\widehat{\beta} \approx \beta$, but there is still uncertainty.

How do we discover the uncertainty in our estimate $\widehat{\beta}$?


# Variance of our Estimate

---

### Variance of OLS Estimator:

- Each element of $\widehat{\beta}_{p \times 1}$ is a random variable
    - each $\widehat{\beta}_k$ has a variance 
    - and a **covariance** with other $\widehat{\beta}_{j\neq k}$
- These variances and covariances are found in the variance-covariance matrix of $\widehat{\beta}$
    - a $p \times p$ matrix
    - Diagonal elements are the **variances** for $\widehat{\beta}_{1} \dots \widehat{\beta}_p$
    - off-diagonal elements are **covariances**; 
    - matrix is **symmetric**

---

Variance-Covariance Matrix

$$\scriptsize{\begin{pmatrix} Var(\widehat{\beta}_1) & Cov(\widehat{\beta}_1,\widehat{\beta}_2) & Cov(\widehat{\beta}_1,\widehat{\beta}_3) &\ldots & Cov(\widehat{\beta}_1,\widehat{\beta}_p) \\
  Cov(\widehat{\beta}_2,\widehat{\beta}_1) & Var(\widehat{\beta}_2) & Cov(\widehat{\beta}_2,\widehat{\beta}_3) & \ldots & Cov(\widehat{\beta}_2,\widehat{\beta}_p) \\ 
  Cov(\widehat{\beta}_3,\widehat{\beta}_1) & Cov(\widehat{\beta}_3,\widehat{\beta}_2) & Var(\widehat{\beta}_3) & \ldots & Cov(\widehat{\beta}_3,\widehat{\beta}_p) \\ 
  \vdots & \vdots & \vdots & \ddots & \vdots \\
  Cov(\widehat{\beta}_p,\widehat{\beta}_1) & Cov(\widehat{\beta}_p,\widehat{\beta}_2) & Cov(\widehat{\beta}_p, \widehat{\beta}_3) & \ldots & Var(\widehat{\beta}_p)\end{pmatrix}}$$


---

### Variance of OLS Estimator:

**How do we use the variance-covariance matrix?**

- The square-root of diagonal elements (variances) gives **standard error** for each parameter estimate in $\widehat{\beta}$ (hypothesis testing)

- The off-diagonal elements can help answer: $\beta_2 + \beta_3 \neq 0$. We need $Cov(\beta_2, \beta_3)$ to get $Var(\beta_2 + \beta_3)$. (complex hypothesis testing, e.g. interaction effects )

---

### Variance-Covariance Matrix

$\widehat{\beta} = \mathbf{\beta} + (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon$, So:

$$cov(\widehat{\beta}|X) = E((\widehat{\beta} - \beta)(\widehat{\beta} - \beta)' | X)$$

$$ = E(  ((\mathbf{X}'\mathbf{X})^{-1} \mathbf{X}'\epsilon)((\mathbf{X}'\mathbf{X})^{-1} \mathbf{X}'\epsilon)' | X)$$

$$ = E(  ((\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\epsilon)(\epsilon'\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}) | X)$$

$$ = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'E(\epsilon\epsilon'|X)\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$

What really matters here is: $E(\epsilon\epsilon'|X)$

---

### Variance-Covariance Matrix

All variance-covariance matrices are "sandwiches":

$$ (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'E(\epsilon\epsilon'|X)\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$

$(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$ is the bread; $E(\epsilon\epsilon'|X)$ is the meat. 

Later on, we can make different choices of "meat".

---

### Variance-Covariance Matrix

In groups: if $\epsilon$ is a vector that is $n \times 1$ with elements $(\epsilon_1 \ldots \epsilon_n)$

1. What are the dimensions of the matrix: $\epsilon\epsilon'$?

2. What are the elements on the **diagonal** of the matrix?

3. What is the **expected value** of the elements on the **off-diagonal** 

Hints: We have assumed that $E(\epsilon) = 0$; $\epsilon_i$ are **independent** of each other

---

### Variance-Covariance Matrix

$$E(\epsilon\epsilon'|X) = \sigma^2\mathbf{I}_{n\times n}$$ 

$\epsilon\epsilon'$ is $n \times n$ matrix with $ij$th elements $\epsilon_i\epsilon_j$

- this is the variance covariance matrix of the **errors** (we don't subtract means because $E(\epsilon) = 0$)

---

### Variance-Covariance Matrix

Taking the expectation:

- If $i \neq j$, the elements of the matrix are $E(\epsilon_i\epsilon_j)$. Because, by assumption, $E(\epsilon) = 0$, and $\epsilon$s are **independent**,  $E(\epsilon_i\epsilon_j) = E(\epsilon_i)E(\epsilon_j) = 0 \times 0 = 0$
- but, for $i = j$, $E[\epsilon_i^2] = E[(\epsilon_i - E[\epsilon])^2]$ (a **variance**)
- and because $\epsilon$ is **identically distributed**, then $Var(\epsilon_i) = Var(\epsilon) = \sigma^2$ for all $\epsilon$

The result is a $n \times n$ matrix, with $\sigma^2$ on the diagonal, $0$ everywhere else.

---

### Variance-Covariance Matrix

$$ = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\sigma^2\mathbf{I}_{n\times n}\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$

$$ = \sigma^2(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}$$
$$cov(\widehat{\beta}|X) = \sigma^2(\mathbf{X}'\mathbf{X})^{-1}$$

This gives us the variance-covariance matrix of $\widehat{\beta}$. Can we start computing standard errors for $\widehat{\beta}$ now?

---

### Variance-Covariance Matrix

We use **residuals** to stand in for $\epsilon$.

$$\widehat{\sigma^2} = \frac{\sum\limits_i^n e_i^2}{n - p}$$


$E[\widehat{\sigma^2} | X] = \sigma^2$ (unbiased)

- Why $n-p$? The variance of the sample mean required $n-1$, because we calculated the mean first, so we lost 1 degree of freedom. With bivariate regression, we lose 2 df. With $p$ variables, we lose $p$ degrees of freedom.


---

### Variance-Covariance Matrix

so, we **estimate** the variance-covariance matrix: 

$\widehat{cov}(\widehat{\beta}) = \widehat{\sigma^2}(\mathbf{X}'\mathbf{X})^{-1}$

---

### Standard Errors of $\widehat{\beta_p}$

The variance ($p$th diagonal element of $\widehat{cov}(\widehat{\beta})$) is

$Var(\hat{\beta_p}) = \frac{\hat{\sigma^2}}{\sum\limits_i^n (X^*_{ip} - \overline{X_p^*})^2}$

So the standard error is:

$SE(\hat{\beta_p}) = \sqrt{Var(\hat{\beta_p})} = \frac{\widehat{\sigma}}{\sqrt{n}\sqrt{Var(X_p^*)}}$

- standard errors get smaller (more precise): with growing $n$, increased (residual) variation in $X_p$, smaller error variance ($\sigma^2$)

---

### Sampling Distribution of $\hat{\beta}$

**asymptotic normality**:

Even if $\epsilon$ not normally distributed: under our assumptions, then $\lim_{n\to\infty}$ sampling distribution of $\hat{\beta}$ is normally distributed by the central limit theorem. Reasonable if $n$ is moderately large and data is not highly skewed. 

Because we **estimate** standard errors: $t$ statistic for hypothesis testing of $\widehat{\beta}$  asymptotically $t$ distributed:

---

### Standard errors

In sum:

- We get $\widehat{\sigma^2}$ by taking variance of the residuals ($e$)
- We take diagonal of variance-covariance matrix to get variances of slopes ($\widehat{\sigma^2}(\mathbf{X}'\mathbf{X})^{-1}$)
- Take square root of variances to standard errors for $\widehat{\beta}$

---

### Standard errors

For standard errors of $\widehat{\beta}$ to be correct, we must assume:

1. $Y$ is generated by $\mathbf{X\beta} + \epsilon$
2. $\epsilon_i$ are independent, identically distributed, with variance $\sigma^2$ for *all* $i$
3. $X_i \perp \!\!\! \perp \epsilon_i$

If **any assumption is wrong**: standard errors will be incorrect.

If assumption 1 and 3 are wrong: unbiasedness of $\widehat{\beta}$ does not hold.

---

In groups:

Download https://pastebin.com/LeSpqKKk

1. Regress `gop_change` on `enlistment` and an intercept using matrix calacualtions
2. Create the residuals `e`
3. Create the variance-covariance matrix
4. Get the coefficient and standard error for `enlistment`
5. For each of the three least squares assumptions:

- what is an example of the assumption being wrong in this case?
- why would it lead standard errors to be wrong?

---

### OLS Assumptions:
 
 When do things go wrong?
 
- We incorrectly specify the model (it is multiplicative, or we forgot a term)
 
>- $Y_i = \alpha + \beta_1 x + \beta_2 x^2 + \epsilon$
 
>- $Y_i = \alpha + \beta_1 x + \beta_2 z + \epsilon$


- Errors are **heteroskedastic**: errors have different variances. 
 
 >- E.g.: At higher levels of education, incomes take on wider range of values (think PhD  vs. MBA students earning potential), so the errors are larger.
 
---

### OLS Assumptions:
 
- Errors are not **independent**: errors for different observations are correlated.
 
>- We observe the same counties over time. The error we observer for county $i$ at time $t$ is likely to be correlated with the error of county $i$ at time $t+1$, because it is the same place.
 
- $\epsilon$ is not independent of $X$. The errors are in fact correlated with $X$:
 
>- This is "confounding". Our estimate $\hat{\beta}$ as the coefficient of $x$ will be biased. If we excluded $z$ from the equation incorrectly, the bias in $x$ will be a function of the relationships between $x$ and $z$, and between $z$ and $y$.
  


---

### Conclusion:

1. $Y$ is generated by $\mathbf{X\beta} + \epsilon$
2. $\epsilon_i$ are independent, identically distributed, with variance $\sigma^2$ for *all* $i$
3. $X_i \perp \!\!\! \perp \epsilon_i$

With these assumptions:

- $\widehat{\beta}$ is **unbiased** estimator for $\beta$
- known sampling distribution of $\widehat{\beta}$


