---
title: "POLI 110: Confounding"
author: "Michael Weaver"
date: "March, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(haven)
require(data.table)
require(ggplot2)
require(magrittr)
require(ggdag)
```

# Confounding


## Introduction

Last lecture, we addressed how we assess causal claims. In particular, we were interested in whether social distancing reduced the spread of disease.

Since then, the BC CDC has said that they estimate that [social distancing has reduced the rate of spread of COVID-19](https://www.cbc.ca/news/canada/british-columbia/physical-distancing-bc-covid19-coronavirus-1.5512269).

## Introduction

But as we noted, there is a **fundamental problem of causal inference**: we don't observe what the infection rate would look like in British Columbia in the absence of social distancing, because that is a **counterfactual**; BC did implement social distancing nearly 3 weeks ago, so we only observe infection rates in the presence of distancing.

We can only <u>estimate</u> the counterfactual:

- The BC CDC uses mathematical models of the disease (we won't be covering this approach) and data they have collected.
- Or we can compare COVID-19 cases in BC with other countries that were slower to impose social distancing (e.g. Italy, United States). This is **correlation**.

## Introduction

**Correlation** can suffer from two sources of error:

**random error**: we observe patterns in $X$ (independent variable) and $Y$ (dependent variable) by chance, when there is in fact no relationship.

**bias (confounding)**: the **observed pattern** between $X$ and $Y$ is not the **true causal relationship** between $X$ and $Y$.

## Confounding

This lecture will discuss in more detail:

- What is confounding?
- Why does it happen?
    - What circumstances make it happen? 
- What is the **direction** of the bias it produces?

# Confounding

## Confounding

In **correlation**, we solve the fundamental problem of causal inference by replacing an **unobservable** counterfactual with a **different, observable** case with a **different exposure to the "cause"** 

So, when we investigated the effects of social distancing on flattening the curve during the Spanish Flu, we compared the two **factual**/**observed** potential outcomes: $\mathrm{Peak \ Mortality \ Rate}_{St. Louis}(\mathrm{More \ Social \ Distance})$ versus $\mathrm{Peak \ Mortality \ Rate}_{Philadelphia}(\mathrm{Less \ Social \ Distance})$.

## Confounding

But the potential outcomes of Philadelphia and St. Louis (marked in $\color{red}{\mathrm{red}}$ ) are **unknown** because they are **counterfactual**. We don't know whether more social distancing caused lower peak mortality rates, because we don't observe these two cities in the alternative universe where they adopted a different social distancing policy.


| City | Less SD | More SD |
|:-----|:----:|:---:|
| Philadelphia | $\mathrm{Peak \ MR_{Phil.}(Less \ SD)}$ | $\color{red}{\mathrm{Peak \ MR_{Phil.}(More \ SD)}}$ |
| | | |
| St. Louis | $\color{red}{\mathrm{Peak \ MR_{St. L.}(Less \ SD)}}$ | $\mathrm{Peak \ MR_{St. L.}(More \ SD)}$ |


## Confounding

Using the correlation between social distancing and peak mortality in these two cities to provide evidence of causality **implicitly** assumes the **observed** Peak Mortality in Philadelphia (St. Louis) is identical with the **counterfactual** Peak Mortality in St. Louis (Philadelphia)

| City | Less SD | More SD |
|:-----|:----:|:---:|
| Philadelphia | $\boxed{\mathrm{Peak \ MR_{Phil.}(Less \ SD)}}$ | $\boxed{\color{red}{\mathrm{Peak \ MR_{Phil.}(More \ SD)}}}$ |
| | $\Downarrow$ | $\Uparrow$ |
| St. Louis | $\boxed{\color{red}{\mathrm{Peak \ MR_{St. L.}(Less \ SD)}}}$ | $\boxed{\mathrm{Peak \ MR_{St. L.}(More \ SD)}}$ |

## Confounding

There are **two ways** of explaining confounding. These two explanations have **identical**, but I introduce both in case one makes more sense than the other.


## Confounding

### **Explanation 1**:

Confounding happens when cases with **different levels of $X$** have different (factual and counterfactual) potential outcomes.

**In our example:**

We had been assuming that in the counterfactual world where Philadelphia imposed more social distancing, the peak of the curve $(\color{red}{\mathrm{Peak \ MR_{Phil.}(More \ SD)}})$ would have been the same as it actually was in St. Louis $(\mathrm{Peak \ MR_{St. L.}(More \ SD)})$ which **did** impose social distancing.(press `p` for notes)

<div class='notes'>
We also had been assuming that in the counterfactual world where St. Louis did not impose more social distancing, the peak of the curve $(\color{red}{\mathrm{Peak \ MR_{St. L.}(Less \ SD)}})$ would have been the same as it actually was in Philadelphia $(\mathrm{Peak \ MR_{Phil.}(Less \ SD)})$ which **did not** impose social distancing.(press `p` for notes)
</div

## Confounding

### **Explanation 1**:

But, what if $\color{red}{\mathrm{Peak \ MR_{Phil.}(More \ SD)}} = \mathrm{Peak \ MR_{Phil.}(Less \ SD)}$: Philadelphia's peak mortality rate would have been the same regardless of whether it had imposed social distancing? And what if $\color{red}{\mathrm{Peak \ MR_{St. L.}(Less \ SD)}} = \mathrm{Peak \ MR_{St. L.}(More \ SD)}$: St. Louis's peak mortality rate would have been the same regardless of whether it had imposed social distancing?

This would imply that social distancing had **no effect** in either city (think about why). But, 
because cities with **more social distancing** (St. Louis) would have had lower peak mortality rates (with or without the social distancing), and cities with **less social distancing** (Philadelphia) would have had higher peak mortality rates with or without the social distancing, we would **observe** that having social distancing reduces peak mortality.

The true causal relationship in this example is $0$; but we **observe** a negative correlation. This is **confounding** or **bias**.

## Confounding

### **Explanation 1**:

This explanation can be a bit tricky to grasp; but the **key** is that the cases that we **observe** with different levels of the social distancing ($X$) have **different** potential outcomes of peak mortality ($Y$): it is **incorrect** to treat the observed outcome of one city as equal to the counterfactual outcome of the other.

This raises a question: **why** do cases with different levels of $X$ have different potential outcomes of $Y$? Why might places that adopt different levels of social distancing have different peak mortality rates, regardless of the public health responses they adopt?

## Confounding

The obvious place to start answering that question is to note: there were **other things that are different** between Philadelhpia and St. Louis in addition to their social distancing policies.

Or, were we to compare COVID-19 in BC to COVID-19 in the New York, we would have to note that **there are many differences** between New York and BC other than their public health responses.

## Confounding

### **Explanation 2**:

Confounding occurs when these differences between cases (captured as variables, e.g. $W$, $Z$, etc.) **affect $X$** and **affect $Y$**. 

The easiest way to understand this is **visually**. 

- We can produce **graphs** of the causal relationships between variables
- We can figure out whether there is going to be **bias**/**confounding** in the correlation of $X$ and $Y$ by reading these graphs. 

## Causal Graphs

These graphs represent a model of the **true causal relationships** between variables.

the **nodes** or **dots** correspond to **variables**

- can be labeled with generic names for independent/dependent variables ($X$, $Y$) or meaningful names (e.g. "Social Distancing", "Peak Mortality")

the **arrows** convey the **direction** of **causality**

- $X \rightarrow Y$ means that $X$ causes changes in $Y$
- $X \leftarrow W$ means that $W$ causes changes in $X$

the **labels** on arrows convey the **direction** of the true **effect**:

- $X \xrightarrow{increase} M$ means that $X$ causes $M$ to **increase**
- $X \xleftarrow{decrease} Z$ means that $Z$ causes $X$ to **decrease**

## Causal Graphs

### **For example**

St. Louis may have been different from Philadelphia because it was exposed to the Spanish Flu later.

Later exposure may have:

- let cities learn from high death rates in other cities, and impose stronger social distancing (later exposure $\xrightarrow{increase}$ social distancing)
- let doctors learn better treatment procedures to save more lives (later exposure $\xrightarrow{decrease}$ peak mortality)


---

press `p` to see notes

```{r, echo = F}
dagify(distance ~ learning_p ,
       deaths ~ learning_m + distance ,
       learning_m  ~ timing,
        learning_p ~ timing,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "deaths" = "Peak\nMortality")
                  ) %>%
  tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', 'decreases', 'increases', 'increases', 'increases', '')),  angle_calc = "along", label_dodge = unit(c(1,-1,1,-1,1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class='notes'>

This graph illustrates the causal relationships described on the preceding slide:

Later Flu exposure $\xrightarrow{increases}$ policy learning $\xrightarrow{increases}$ social distancing

Later Flu exposure $\xrightarrow{increases}$ medical learning $\xrightarrow{decreases}$ peak mortality

This graph also says that: social distancing $\xrightarrow{decreases}$ peak mortality

</div>

## Causal Graphs

We can see **confounding** or **bias** on causal graphs:

If we want to assess whether $X$ causes $Y$ using correlation, that correlation will suffer from **confounding** if, on the causal graph, there is a **backdoor path** from $X$ to $Y$. 

- A "frontdoor" or **causal path** from $X$ to $Y$: it is a chain of one or more arrows that **starts** at $X$ and **points** toward $Y$: $X \rightarrow W \rightarrow Z \rightarrow Y$. Because the direction of causality follows the direction of the arrows, this path means that $X$ causes some change in $Y$.
- A **backdoor** path is a **non-causal** path from $X$ to $Y$: it is a chain of **two** or more arrows that **starts** at $X$ and follows arrows **backwards** out of $X$, then changes direction **once** and follows arrows **toward** $Y$: $X \leftarrow W \leftarrow Z \rightarrow Y$ 

With frontdoor paths and backdoor paths, we expect to see a relationship between $X$ and $Y$, but only in a frontdoor path did $X$ **cause** $Y$. In a backdoor path, $X$ and $Y$ are both **caused** by some other factor, producing an observed correlation that is different from the true causal link between $X$ and $Y$.


---

press `p` for notes

```{r, echo = F}
p = dagify(distance ~ learning_p ,
       deaths ~ learning_m + distance ,
       learning_m  ~ timing,
        learning_p ~ timing,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "deaths" = "Peak\nMortality")
                  )  %>%
  ggdag_paths(layout = 'tree', text = F, text_col = 'black', shadow = T, node = F) + theme_dag() +
  geom_dag_text_repel(force = 0, aes(label = label)) +
  theme(legend.position = 'none')

p$data[13, c("direction", "to", "xend", "yend")] = rep(NA, 4)

p
```

<div class='notes'>

In panel 1 (on the left), we see a causal path: there is an arrow starting at Social distancing  $\xrightarrow{\mathrm{pointing \ toward}}$ Peak Mortality.

</div>


---

press `p` for notes

```{r, echo = F}
p = dagify(distance ~ learning_p ,
       deaths ~ learning_m + distance ,
       learning_m  ~ timing,
        learning_p ~ timing,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "deaths" = "Peak\nMortality")
                  )  %>%
  ggdag_paths(layout = 'tree', text = F, text_col = 'black', shadow = T, node = F) + theme_dag() +
  geom_dag_text_repel(force = 0, aes(label = label)) +
  theme(legend.position = 'none')

p$data[13, c("direction", "to", "xend", "yend")] = rep(NA, 4)

p
```

<div class='notes'>

In panel 2 we see a **backdoor path**: there is an arrow coming **backward** out of Social distancing $\xleftarrow{{\textrm{pointing from}}}$ Policy learning $\xleftarrow{{\textrm{pointing from}}}$ Later Flu exposure. Then, the path changes directions (just once): an arrow comes from later flu exposure $\xrightarrow{\textrm{pointing toward}}$ medical learning $\xrightarrow{\textrm{pointing toward}}$ peak mortality.

This means there is confounding (by explanation 2): something that can differ between cases (the timing of the flu epidemic) affects the $X$ (social distancing) **and** the $Y$ (peak mortality) 
</div>

## Causal Graphs: Confounding

In reality, we don't really **know** the variables and paths on these causal graphs.

Instead, these causal graphs help us think about **possible scenarios** that might produce **bias**/**confounding** of the correlation between $X$ and $Y$.

## Causal Graphs: Confounding

Now it is your turn to practice:

> **Does wearing a surgical mask during everyday activities during cold and flu season reduce a person's risk of contracting a cold or flu?**

We observe this correlation:  **people who wear masks have lower rates of cold and flu infection**

Draw a causal graph where this correlation suffers from **confounding**/**bias**


## Causal Graphs: Confounding

Draw your graphs!

---

Here is one **possible** graph: press `p` to see notes

```{r, echo = F}
dagify(mask ~ ocd + immuno,
       flu ~ hands,
       hands ~ ocd + immuno,
       exposure = "mask", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', 'increases','increases',  'increases', 'increases', '', '')),  angle_calc = "along", label_dodge = unit(c(-1, -1,1,1, 1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class='notes'>

In this causal graph: 

Being **immuno-compromised** $\xrightarrow{increases}$ hand washing and $\xrightarrow{increases}$ mask wearing.

Having **OCD** $\xrightarrow{increases}$ hand washing and $\xrightarrow{increases}$ mask wearing.

Hand washing $\xrightarrow{decreases}$ likelihood of getting the flu.

Mask wearing has **no effect** on getting the flu mask wearing. (no arrow)

</div>

--- 

press `p` for notes

```{r, echo = F}
dagify(mask ~ ocd + immuno,
       flu ~ hands,
       hands ~ ocd + immuno,
       exposure = "mask", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu")
                  )  %>%
  ggdag_paths(layout = 'auto', text = F, text_col = 'black', shadow = T, node = F) + theme_dag() +
  geom_dag_text_repel(force = 0, aes(label = label)) +
  theme(legend.position = 'none')

```


<div class='notes'>

In this causal graph, we can see there are **no** causal paths from mask wearing to getting the flu. But there are two **backdoor paths**.

Panel 1 (Left): We can see that there is a path coming backward out of mask wearing $\xleftarrow{\textrm{pointing from}}$ Immuno-compromised (presumably because people who have a compromised immune system want to be extra safe and avoid getting the flu). Then the path changes directions, with an arrow coming from immuno-compromised and $\xrightarrow{\textrm{pointing toward}}$ hand washing and $\xrightarrow{\textrm{pointing toward}}$ getting the flu. The immuno-compromised also wash their hands more (to avoid illness) and hand washing affects getting the flu.


</div>

--- 

press `p` for notes

```{r, echo = F}
dagify(mask ~ ocd + immuno,
       flu ~ hands,
       hands ~ ocd + immuno,
       exposure = "mask", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu")
                  )  %>%
  ggdag_paths(layout = 'auto', text = F, text_col = 'black', shadow = T, node = F) + theme_dag() +
  geom_dag_text_repel(force = 0, aes(label = label)) +
  theme(legend.position = 'none')

```


<div class='notes'>

In this causal graph, we can see there are **no** causal paths from mask wearing to getting the flu. But there are two **backdoor paths**.

Panel 2 (Right): We can see that there is a path coming backward out of mask wearing $\xleftarrow{\textrm{pointing from}}$ OCD (presumably because people who are OCD fear germs and wear masks to ease anxiety). Then the path changes directions, with an arrow coming from OCD and $\xrightarrow{\textrm{pointing toward}}$ hand washing and $\xrightarrow{\textrm{pointing toward}}$ getting the flu. Those with OCD wash their hands more (also to ease anxiety about germs) and hand washing affects getting the flu.

In both cases, these backdoor paths mean that some other factor (OCD, Immuno-compromised) affects both **mask-wearing** and **getting the flu**. This is confounding by explanation 2.
</div>


## Confounding:
 
These examples illustrate the possibility that if causal graphs include variables **in addition** to the independent and dependent variables, there is a risk of confounding or bias.

Do **all** additional variables produce **confounding**?

**No...** We will discuss three different patterns of variables: some of which have confounding, some which do not.

## Additional Variables: Patterns

- **antecedent variables**
    - sometimes confounding
    - sometimes no confounding
- **intervening variables**
    - no confounding
- **reverse causality**
    - yes, confounding.
      
    
## Antecedent Variables

      
**antecedent variable**: a variable that **affects** $X$

- e.g. in this path, $Z \xrightarrow{} X \xrightarrow{} Y$, $W$ is an antecedent variable.

- antecedent variables ($Z$) **do not** produce confounding if the **only causal path** from $Z$ to $Y$ passes through $X$.
- antecedent variables **do** produce confounding if there is another **causal path** from $Z$ to $Y$ that does **NOT** include $X$.


## Antecedent Variable: No Confounding

press `p` for notes

```{r, echo = F}

dagify(
       flu ~ hands,
       hands ~ immuno ,
       exposure = "hands", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes Hands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu")
                  )  %>%
  ggdag_classic(layout = 'auto', text_label = 'label', size = 5) + theme_dag() +
  theme(legend.position = 'none')

```

<div class="notes">

In this example, if we are interested in the effect of washing hands ($X$) on getting the flu ($Y$), immuno-compromised is an **antecedent variable**, because it causes hand washing.

However, there is **no confounding** because there is no other **causal path** from being immuno-compromised to getting the flu $Y$. The only path passes through $X$.

</div>

## Antecedent Variable: No confounding

press `p` for notes


```{r, echo = F}
dagify(mask ~ immuno,
       flu ~ hands,
       hands ~ immuno,
       exposure = "mask", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link() +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

In this example, if we are interested in the effect of washing hands ($X$) on getting the flu ($Y$),  being Immuno-compromised is **antecedent variableis ans**, because it causes hand washing.

However, there is **no confounding** because there is no other **causal path** from being immuno-compromised to getting the flu $Y$. Being Immuno-compromised  also causes mask-wearing; but the graph shows no effect of mask-wearing on getting the flu $Y$. So the only causal paths from being Immuno-Compromised to getting the flu ($Y$) passes through hand washing ($X$). 

</div>

## Antecedent Variable: Confounding

press `p` for notes

```{r, echo = F}
dagify(mask ~ immuno,
       flu ~ hands + distance,
       hands ~ immuno,
       distance ~ immuno,
       exposure = "hands", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "ocd" = "OCD",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu",
                  "distance" = "Social\nDistancing")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link() +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

In this example, if we are interested in the effect of washing hands ($X$) on getting the flu ($Y$),being Immuno-compromised is an **antecedent variable**, because it causes hand washing.

In this case, there is **confounding!** There is a **causal path** from being Immuno-compromised $\xrightarrow{\textrm{pointing toward}}$ social distancing $\xrightarrow{\textrm{pointing toward}}$ getting the flu. Because there is now **causal path** from the antecedent variable (being immuno-compromised) to the dependent variable (getting the flu) that **excludes** the independent variable (hand washing), there is now a **backdoor path** between $X$ (hand washing) and $Y$ (getting the flu) that passes through the antecedent variable (being immuno compromised).

</div>

## Antecedent Variable: No Confounding

press `p` for notes

```{r, echo = F}
dagify(mask ~ immuno + smog,
       flu ~ hands + indoors,
       hands ~ immuno,
       indoors ~ smog,
       exposure = "hands", 
       outcome = 'flu',
       labels = c("mask" = "Wears\nMask",
                  "hands" = "Washes\nHands",
                  "smog" = "Smoggy\nCity",
                  "immuno" = "Immuno-\nCompromised",
                  "flu" = "Get Flu",
                  "indoors" = "Stays\nIndoors")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link() +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

In this example, if we are interested in the effect of washing hands ($X$) on getting the flu ($Y$), being Immuno-compromised is an **antecedent variable**, because it causes hand washing.

In this causal graph, there is **no confounding**! You might think that there is. There **seems to be** a **causal path** from being Immuno-compromised $\xrightarrow{\textrm{pointing toward}}$ mask wearing $\xleftarrow{\textrm{pointing from}}$ smog $\xrightarrow{\textrm{pointing toward}}$ staying indoors $\xrightarrow{\textrm{pointing toward}}$ getting the flu. This **is** a path from being immuno-compromised (the antecedent variable) to the getting the flue ($Y$) that excludes the hand washing ($X$).

But look more closely, it is not a **causal path**: being immuno-compromised $\xrightarrow{causes}$ mask wearing; living in a smoggy city also $\xrightarrow{causes}$ mask wearing. But mask wearing does not cause living in a smoggy city. Going backwards along the arrow $\textrm{wears mask} \xleftarrow{} \textrm{Smoggy city}$ is a **non-causal** step.

Another way of seeing this is: there is no **backdoor path** from hand washing to getting the flu, because there arrows would have to change direction **twice**: hand washing $\xleftarrow{\textrm{backwards toward}}$ being immuno-compromised $\xrightarrow{\textrm{forwards toward}}$ mask wearing $\xleftarrow{\textrm{backwards toward}}$ smoggy city.


Because there is now **causal path** from the antecedent variable (being immuno-compromised) to the dependent variable (getting the flu) that **excludes** the independent variable (hand washing), there is now a **backdoor path** between $X$ (hand washing) and $Y$ (getting the flu) that passes through the antecedent variable (being immuno compromised).

</div>

## Intervening Variables

**intervening variable**: a variable that **affects** $Y$ and is **affected by** $X$. 

- e.g. in this path, $X \xrightarrow{} W \xrightarrow{} Y$, $W$ is an intervening variable.

- intervening variables ($W$) **do not** produce confounding because they are on the  **causal path** from $X$ to $Y$. They do not produce backdoor path.

## Intervening Variable

press `p` to see notes

```{r, echo = F}
dagify(
       deaths ~ transmission ,
       transmission  ~ distance,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "transmission" = "Disease\nTransmission",
                  "deaths" = "Peak\nMortality")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', 'increases', '')),  angle_calc = "along", label_dodge = unit(c(-1,-1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

In this example, if we are interested in the effect of social distancing ($X$) on peak mortality ($Y$), the disease transmission is an intervening variable.

Here we can see there is no **backdoor path** from social distancing to peak mortality: there is only a causal path and the intervening variable falls along this causal path. Thus, there is no confounding.

</div>

## Reverse Causality

**reverse causality** describes the situation where the **dependent variable** $Y$ actually causes the **independent variable** $X$. 

So while we use the correlation to describe the effect of $X$ on $Y$: $X \to Y$, the correlation in fact is the result of the effect of $Y$ on $X$: $Y \to X$. 

This is a special case of **bias** or **confounding**.

---

|  | Third Variable? | Key Attribute | Confounding? |
|--------------------------|-----------------|---|----------------------------|
| Antecedent Variables ($Z$) | Yes | $Z \to X$ | If only **causal path** from $Z$ to $Y$ contains $X$: No <br> If a **causal path** from $Z$ to $Y$ excludes $X$: Yes |
| Intervening Variables ($W$) | Yes | $X \to W \to Y$ | No |
| Reverse Causality | No  | $Y \to X$ | Yes |


# Direction of Bias

## Direction of Bias

**Confounding** or **bias** in correlations means that the correlations we observe are systematically different from the **true causal relationship** between $X$ and $Y$.

If there is confounding, we need to be able to reason: 

- is the correlation that we observe **more positive** than the **true effect** (**upward bias**)?
- is the correlation that we observe **more negative** than the **true effect** (**downward bias**)?

## Direction of Effects

Before we can read these graphs to find whether bias is **upward** or **downward**, we need to be able to read causal graphs to find the direction of the effect for a causal path.

To do this, we first convert the words along the arrows on a causal path into numbers: "increases" into $+1$ and "decreases" into $-1$. Then, we take the **product**  of these $+1$s and $-1$s (multiply all of them together). 

- If the **product** is $+1$, then the overall effect is positive
- If the **product** is $-1$, then the overall effect is negative

## Direction of Effect


press `p` to see notes

```{r, echo = F}
dagify(
       deaths ~ distance,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "transmission" = "Disease\nTransmission",
                  "deaths" = "Peak\nMortality")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', '')),  angle_calc = "along", label_dodge = unit(c(-1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

This is the simplest case: there is only Social Distancing ($X$) and Peak Mortality ($Y$). Here we see that Social Distancing **decreases** Peak Mortality. We convert "decreases" to $-1$. There is nothing to multiply it by, so the total effect along this path ($\textrm{Social Distancing} \to \textrm{Peak Mortality}$) is negative (more social distancing reduces peak mortality).


</div>

## Direction of Effect

press `p` to see notes

```{r, echo = F}
dagify(
       deaths ~ transmission ,
       transmission  ~ distance,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "transmission" = "Disease\nTransmission",
                  "deaths" = "Peak\nMortality")
                  ) %>%
  tidy_dagitty(layout='auto') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', 'increases', '')),  angle_calc = "along", label_dodge = unit(c(-1,-1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class="notes">

This is more complicated: we see that disease tranmission is on the path between Social Distancing ($X$) and Peak Mortality ($Y$). Here we see that Social Distancing **decreases** Disease Transmission (because there are fewer opportunities for people to spread it to others). We convert "decreases" to $-1$. Disease Transmission **increases** Peak Mortality (because as more people get the disease more rapidly, more people will potentially die). We convert "increases" into $+1$. 

Then we take the product of these two numbers: $1 \times -1 = -1$. Because the overall product is $-1$ the total effect of $X$ on $Y$ along this path ($\textrm{Social Distancing} \to \textrm{Disease Transmission} \to \textrm{Peak Mortality}$) is negative (more social distancing reduces peak mortality). This makes sense: greater disease transmission **increases** mortality, and social distancing **reduces** disease transmission. So it must **reduce** mortality. 

Even for much longer paths, taking the product will give us the right answer.

</div>

## Confounding: Direction of Bias

In order for there to be **confounding** there must be an (antecedent) variable $Z$ that has a **causal path** to $X$  that also has  **another causal path**  toward $Y$.

To find the **direction of the bias**, we simply use the rules above to get the direction of the effect for both of these paths:

- $Z \to \dots \to X$
- $Z \to \dots \to Y$

Then take the **product** of those two paths.

## Confounding: Direction of Bias

Product of **signs** on relationships between $W \to X$, $W \to Y$ gives us <u>**direction**</u> of **bias** created by **confounding**

|  | $W \xrightarrow{+} X$ | $W \xrightarrow{-} X$ |
|-----------------------|-------------------------------------|--------------------------------|
| $W \xrightarrow{+} Y$ | $Correlation(X,Y)$<br>Upward Bias (+) | $Correlation(X,Y)$<br>Downward Bias (-) |
| $W \xrightarrow{-} Y$ | $Correlation(X,Y)$<br>Downward Bias (-) | $Correlation(X,Y)$<br>Upward Bias (+) |

## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Decreases', 'Increases', 'Bias \n(?)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```

## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Decreases', 'Increases', 'Bias \n(Downward)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```

## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Decreases', 'Decreases', 'Bias \n(?)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```


## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Decreases', 'Decreases', 'Bias \n(Upward)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```

## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Increases', 'Increases', 'Bias \n(?)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```


## Confounding: Direction of Bias

```{r, echo = F}
dagify(x ~ w,
       y ~ w,
       y ~~ x,
       exposure = "x", 
       outcome = 'y'
       ) %>%
tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_node() +
  geom_dag_edges_link(mapping = aes(label = c('Increases', 'Increases', 'Bias \n(Upward)', '')), 
                      angle_calc = "along", 
                      label_dodge = unit(c(-1,1,-1),'lines'), 
                      edge_linetype = rep(c(1,1,3), each = 100),
                      arrow = grid::arrow(length=grid::unit(c(10,10,0), 'pt'), type = 'closed')) +
  geom_dag_text(colour = 'white') +
  theme_dag() +
  scale_adjusted()
```

---

What is the direction of bias for correlation between $X$ and $Y$ here? press `p` to see notes

```{r, echo = F}
dagify(distance ~ learning_p ,
       deaths ~ learning_m + distance ,
       learning_m  ~ timing,
        learning_p ~ timing,
       exposure = "distance", 
       outcome = 'deaths',
       labels = c("distance" = "(X) Social\nDistancing",
                  "learning_p" = "Policy\nLearning",
                  "learning_m" = "Medical\nLearning",
                  "timing" = "Later Flu",
                  "deaths" = "(Y) Peak\nMortality")
                  ) %>%
  tidy_dagitty(layout='tree') %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(mapping = aes(label = c('decreases', 'decreases', 'increases', 'increases', 'increases', '')),  angle_calc = "along", label_dodge = unit(c(1,-1,1,-1,1),'lines')) +
  geom_dag_text(mapping = aes(label = label), colour = 'black') +
  theme_dag() +
  scale_adjusted()

```

<div class='notes'>

If we want to evaluate the effect of **social distancing** ($X$) on peak mortality ($Y$) using correlation, there will be confounding, because there is a backdoor path from social distancing to peak mortality.

To get the direction of the this bias:

1. First, get the direction of the effect of Later Flu on social distancing: $\textrm{Later Flu} \xrightarrow{increases} \textrm{Policy Learning} \xrightarrow{increases} \textrm{Social Distancing}$ implies $+1 \times +1 = 1$
2. Second, get the direction of the effect of later flu on peak mortality: $\textrm{Later Flu} \xrightarrow{increases} \textrm{Medical Learning} \xrightarrow{decreases} \textrm{Peak Mortality}$ implies $+1 \times -1 = -1$
3. Then take the product of the directions for both paths; $+1 \times -1 = -1$

This means that the confounding generates a **downward bias**: even though this causal graph shows that social distancing decreases peak mortality, the correlation we would observe would be even **more** negative relationship than in truth.
</div>